---
title: Loss estimation
author: Blaž Erzar
output: pdf_document
---

In practice, many machine learning practitioners evaluate their models using
train-test splits or cross-validation. When using these approaches, we are
only estimating the true risk of the model. In this report, we present
findings on how the estimated model risk compares to the true risk proxy in
different scenarios.

```{r include = FALSE}
library(ggplot2)
library(dplyr)

toy_data <- function(n, seed = NULL) {
  set.seed(seed)
  x <- matrix(rnorm(8 * n), ncol = 8)
  z <- 0.4 * x[, 1] - 0.5 * x[, 2] + 1.75 * x[, 3] - 0.2 * x[, 4] + x[, 5]
  y <- runif(n) > 1 / (1 + exp(-z))
  return(data.frame(x = x, y = y))
}

log_loss <- function(y, p) {
  -(y * log(p) + (1 - y) * log(1 - p))
}
```

# A proxy for the true risk

```{r include = FALSE}
df_dgp <- toy_data(100000, 1)
```

The model's risk converges to the true risk in probability. Because it is an
average, it is distributed normally around the true value. Its variance drops
with the sample size. To compute the variance of the log-loss, we train a
model and compute the log-loss for every data instance. The standard error,
i.e., the standard deviation of the estimator, is then the standard deviation
of these log-losses divided by the square root of the sample size. We can see
that the standard deviation of the log-losses is reduced to the third decimal
digit.

```{r}
h_dgp <- glm(y ~ ., data = df_dgp, family = "binomial")
losses <- log_loss(df_dgp$y, predict(h_dgp, df_dgp, type = "response"))
cat(sprintf("Model's risk SE: %.4f\n", sd(losses) / sqrt(nrow(df_dgp))))
```

# Model loss estimator variability due to test data variability

First, we take a look at the variability of the model's risk computed on a
separate test set, which is generated from the same DGP as the training data.
In practice, we usually do not have access to such test sets.

```{r include = FALSE}
data_train <- toy_data(50, 0)
h <- glm(y ~ ., data = data_train, family = "binomial")

# Compute the true model risk from the DGP
compute_model_risk <- function(model, data, aggregate = TRUE) {
  losses <- log_loss(
    data$y,
    predict(model, select(data, !"y"), type = "response")
  )
  if (aggregate) mean(losses) else losses
}
true_risk <- compute_model_risk(h, df_dgp)
```

```{r results = "hold", include = FALSE}
standard_errors <- vector("numeric", 1000)
ci_include <- vector("logical", 1000)
risk_differences <- vector("numeric", 1000)

# Compute CI using empirical risk and check if true risk is contained
for (i in 1:1000) {
  data_test <- toy_data(50, i + 1)
  losses <- compute_model_risk(h, data_test, aggregate = FALSE)
  est_risk <- mean(losses)
  standard_errors[i] <- sd(losses) / sqrt(nrow(data_test))

  ci <- est_risk + c(-1, 1) * 1.96 * standard_errors[i]
  ci_include[i] <- true_risk >= ci[1] && true_risk <= ci[2]

  risk_differences[i] <- est_risk - true_risk
}

baseline_50_50 <- mean(log_loss(df_dgp$y, 0.5))
```

```{r fig.width = 2, fig.height = 2, echo = FALSE}
data.frame(risk_differences) |>
  ggplot(aes(x = risk_differences)) +
  geom_density() +
  xlab(bquote(R[n] - R)) +
  ylab("Density")

cat(
  sprintf("True risk proxy: %.4f\n", true_risk),
  sprintf("Mean difference: %.4f\n", mean(risk_differences)),
  sprintf("0.5-0.5 baseline true risk: %.4f\n", baseline_50_50),
  sprintf("Median standard error: %.4f\n", median(standard_errors)),
  sprintf(
    "Percentage of 95CI that contain the true risk proxy: %.1f%%\n",
    100 * mean(ci_include)
  ), sep = ""
)
```

Since the baseline true risk is higher than the model's true risk, the model
successfully learned patterns from the data.

As expected by the convergence of the model's risk to the true risk, the
difference is distributed with a mean close to zero. This means that the
risk estimate using a test set is unbiased and a sensible estimate in practice
if we have access to a test set from our DGP.

It can be seen from the standard error, that there is some variability in the
estimate of the risk. This is expected because a smaller test set with 50
instances will not have the same distribution as the DGP. Since the estimate
has some uncertainty, it should be reported in practice.

We also observe that the true risk is contained less than 95% of the time,
meaning the standard error is slightly underestimated.

With a bigger train set, the true risk, as well as the estimate, would probably
decrease, because the model would have more data to learn from. A bigger train
set would probably also decrease the variance of the estimate because all
such sets would be more representative of the DGP. If the train set was too
small, the model would not be able to learn anything and could perform even
worse than the baseline.

With a bigger test set, the variance of the difference would be smaller,
because the test set would be more representative of the DGP. Similarly, for a
smaller test set, the variance would be bigger.

# Overestimation of the deployed model's risk

In practice, we usually deploy models that are trained on a dataset that is
larger than the train set used to estimate the model's performance. Here, we
compare the true risk of two models, $h_1$ trained on 50 observations and $h_2$
trained on an additional 50 observations. Both train sets have 50 observations
in common.

```{r warning = FALSE, include = FALSE}
risk_differences <- vector("numeric", 50)

for (i in 1:50) {
  toy_1 <- toy_data(50, 2 * i)
  toy_2 <- toy_data(50, 2 * i + 1)
  models <- lapply(list(toy_1, rbind(toy_1, toy_2)), function(data) {
    glm(y ~ ., data = data, family = "binomial")
  })
  risks <- sapply(models, function(h) compute_model_risk(h, df_dgp))
  risk_differences[i] <- risks[1] - risks[2]
}
```

```{r echo = FALSE}
cat("Summary of true risk h1 - true risk h2:\n")
summary(risk_differences)
```

As expected, the model trained on a larger dataset performs better in general.
The risk of $h_2$ is usually lower than the risk of $h_1$, meaning their
difference is positive.

In practice, we should keep in mind that the risk estimation is computed on a
different model than the one that is deployed. After deployment, we could
additionally evaluate the model on a separate test set, perhaps even in
parallel with the already deployed model, if it exists.

If the train sets were smaller, the difference would probably be even bigger,
because each additional observation would have a larger impact. This is
especially true because the models might not converge with such a small train
set.

If the train sets were bigger, the difference would be smaller, because the
smaller train set for $h_1$ would already be representative of the DGP and
additional observations would not have a big impact.

# Loss estimator variability due to split variability

Most commonly, we do not have access to an additional test set from the DGP,
so we split the data into train and test sets.

```{r results = "hold", warning = FALSE, include = FALSE}
data <- toy_data(100, 0)
h_0 <- glm(y ~ ., data = data, family = "binomial")
true_risk <- compute_model_risk(h_0, df_dgp)

standard_errors <- vector("numeric", 1000)
ci_include <- vector("logical", 1000)
risk_differences <- vector("numeric", 1000)

for (i in 1:1000) {
  train_indices <- sample(nrow(data), nrow(data) / 2)
  train <- data[train_indices, ]
  test <- data[-train_indices, ]
  h <- glm(y ~ ., data = train, family = "binomial")

  losses <- compute_model_risk(h, test, aggregate = FALSE)
  est_risk <- mean(losses)
  standard_errors[i] <- sd(losses) / sqrt(nrow(test))

  ci <- est_risk + c(-1, 1) * 1.96 * standard_errors[i]
  ci_include[i] <- true_risk >= ci[1] && true_risk <= ci[2]

  risk_differences[i] <- est_risk - true_risk
}
```

```{r fig.width = 2, fig.height = 1.8, echo = FALSE}
data.frame(risk_differences) |>
  ggplot(aes(x = risk_differences)) +
  geom_density() +
  xlab(bquote(R[n] - R)) +
  ylab("Density")

cat(
  sprintf("True risk proxy: %.4f\n", true_risk),
  sprintf("Mean difference: %.4f\n", mean(risk_differences)),
  sprintf("Median standard error: %.4f\n", median(standard_errors)),
  sprintf(
    "Percentage of 95CI that contain the true risk proxy: %.1f%%\n",
    100 * mean(ci_include)
  ), sep = ""
)
```

As expected, this estimate is highly biased. More accurately, it is positively
biased and pessimistic. This means that the estimated risk is higher than the
true risk. Also, the standard error is slightly higher than the standard error
of separate test set estimates, but this error is even more underestimated.
This can be seen from the coverage of the 95% confidence interval, which is
much lower here. In each split, we get a variability of the model performance
due to different train sets and a variability of the estimate due to different
test sets.

There are also instances, where the estimated risk is much higher than the true
risk. This is visible from the long right tail of the density plot. This
probably occurs when the train set is not representative of the DGP, and the
model cannot properly learn and performs much worse.

For practical purposes, we should note that train-test split estimation is
highly biased, especially for small datasets, where a smaller train set causes
a large decrease in the model's performance. Even if we report the uncertainty,
it will usually be underestimated.

If the dataset was larger, the bias would probably be smaller, because the
train set would be more representative of the DGP and the difference with the
original train set would be smaller. The estimate would still have some
variance, but it would be smaller if the dataset was large enough for its
splits to be representative. Both the estimated and the true risk would be
lower because the model would have more data to learn from.

If the proportion of the train set was bigger, the model would perform better
and more similarly to the model trained on the whole dataset. This means that
the bias would be smaller, but because the test set would be smaller, the
variance would increase. On the other hand, a smaller proportion of the train
set would increase the bias, but decrease the variance.

# Cross-validation

Instead of train-test split, cross-validation is used many times. We compare
cross-validation with a different number of folds and repetitions, while also
drawing a new dataset from the DGP in each iteration.

```{r include = FALSE}
cv_estimate <- function(data, folds, repeats = 1, seed = NULL) {
  set.seed(seed)
  losses <- matrix(NA, nrow = repeats, ncol = nrow(data))

  for (i in 1:repeats) {
    # Create folds
    folds_mask <- sample(rep(1:folds, each = nrow(data) / folds))

    for (j in 1:folds) {
      test_mask <- folds_mask == j
      train <- data[folds_mask != j, ]
      test <- data[test_mask, ]

      # Train on all other folds, test on the current fold
      h <- glm(y ~ ., data = train, family = "binomial")
      losses[i, test_mask] <- compute_model_risk(h, test, aggregate = FALSE)
    }
  }

  return(colMeans(losses))
}
```

```{r cache = TRUE, warning = FALSE, include = FALSE}
folds_repeats <- matrix(c(2, 4, 10, 10, 100, 1, 1, 1, 20, 1), nrow = 5)
cv_names <- c("2-fold", "4-fold", "10-fold", "10-fold-20-rep", "loocv")

standard_errors <- matrix(NA, nrow = 5, ncol = 500)
ci_include <- matrix(NA, nrow = 5, ncol = 500)
risk_differences <- matrix(NA, nrow = 5, ncol = 500)

for (i in 1:500) {
  data <- toy_data(100, i + 1)
  h_0 <- glm(y ~ ., data = data, family = "binomial")
  true_risk <- compute_model_risk(h_0, df_dgp)

  for (experiment in 1:5) {
    # Compute CV estimate
    folds <- folds_repeats[experiment, 1]
    repeats <- folds_repeats[experiment, 2]
    losses <- cv_estimate(data, folds, repeats, seed = i)

    est_risk <- mean(losses)
    standard_errors[experiment, i] <- sd(losses) / sqrt(nrow(data))

    ci <- est_risk + c(-1, 1) * 1.96 * standard_errors[experiment, i]
    ci_include[experiment, i] <- true_risk >= ci[1] && true_risk <= ci[2]

    risk_differences[experiment, i] <- est_risk - true_risk
  }
}
```

```{r results = "hold", fig.height = 3.8, echo = FALSE}
data.frame(
  risk_differences = c(t(risk_differences)),
  method = rep(cv_names, each = 500)
) |>
  ggplot(aes(x = risk_differences)) +
  geom_density() +
  facet_wrap(~ factor(method, cv_names), ncol = 4) +
  xlab(bquote(R[n] - R)) +
  ylab("Density")
```

We can observe that increasing the number of folds decreases the bias of the
estimate. This is expected because the train set is getting bigger and more
similar to the original dataset. This also means that the variance gets smaller
with the number of folds because the model performance is more stable. However,
drawing a new dataset in each iteration has caused the uncertainty to be more
underestimated when compared to the previous task.

The smallest bias is achieved with the leave-one-out cross-validation, while
10-fold, 10-fold with 20 repetitions and LOO have similar variances. Out of
these, 10-fold cross-validation with 20 repetitions underestimated the
uncertainty the least.

From these results, we can conclude that using 2-fold cross-validation is
not the best option in practice. Usually, LOO performs the best, but we can
achieve similar results with 10-fold cross-validation with 20 repetitions and
save a lot of computational time.

```{r results = "hold", echo = FALSE}
for (i in 1:5) {
  cat(
    cv_names[i], ":\n",
    sprintf("  Mean difference: %.4f\n", mean(risk_differences[i, ])),
    sprintf("  Median standard error: %.4f\n", median(standard_errors[i, ])),
    sprintf(
      "  Percentage of 95CI that contain the true risk proxy: %.1f%%\n",
      100 * mean(ci_include[i, ])
    ), sep = ""
  )
}
```

\newpage

# A different scenario

Increasing the size of the dataset from 100 to 1000, we can observe that the
results are much different from the previous task. All methods perform very
similarly, there being negligible differences between 10-fold, 10-fold with 20
repetitions and LOO. The 2-fold and 4-fold cross-validations perform only
slightly worse, which might not even be significant.

The reason for this is the fact that the dataset is much bigger. Even the
folds in the 2-fold cross-validation that had a high variance before, are now
more representative of the DGP than the whole previous dataset was. We can also
observe that this is the only task where the uncertainty is not underestimated,
in some cases even being slightly overestimated.

```{r cache = TRUE, warning = FALSE, include = FALSE}
folds_repeats <- matrix(c(2, 4, 10, 10, 100, 1, 1, 1, 20, 1), nrow = 5)
cv_names <- c("2-fold", "4-fold", "10-fold", "10-fold-20-rep", "loocv")

standard_errors <- matrix(NA, nrow = 5, ncol = 500)
ci_include <- matrix(NA, nrow = 5, ncol = 500)
risk_differences <- matrix(NA, nrow = 5, ncol = 500)

for (i in 1:500) {
  data <- toy_data(1000, i + 1)
  h_0 <- glm(y ~ ., data = data, family = "binomial")
  true_risk <- compute_model_risk(h_0, df_dgp)

  for (experiment in 1:5) {
    # Compute CV estimate
    folds <- folds_repeats[experiment, 1]
    repeats <- folds_repeats[experiment, 2]
    losses <- cv_estimate(data, folds, repeats, seed = i)

    est_risk <- mean(losses)
    standard_errors[experiment, i] <- sd(losses) / sqrt(nrow(data))

    ci <- est_risk + c(-1, 1) * 1.96 * standard_errors[experiment, i]
    ci_include[experiment, i] <- true_risk >= ci[1] && true_risk <= ci[2]

    risk_differences[experiment, i] <- est_risk - true_risk
  }
}
```

```{r results = "hold", fig.height = 3.8, echo = FALSE}
data.frame(
  risk_differences = c(t(risk_differences)),
  method = rep(cv_names, each = 500)
) |>
  ggplot(aes(x = risk_differences)) +
  geom_density() +
  facet_wrap(~ factor(method, cv_names), ncol = 4) +
  xlab(bquote(R[n] - R)) +
  ylab("Density")
```

```{r results = "hold", echo = FALSE}
for (i in 1:5) {
  cat(
    cv_names[i], ":\n",
    sprintf("  Mean difference: %.4f\n", mean(risk_differences[i, ])),
    sprintf("  Median standard error: %.4f\n", median(standard_errors[i, ])),
    sprintf(
      "  Percentage of 95CI that contain the true risk proxy: %.1f%%\n",
      100 * mean(ci_include[i, ])
    ), sep = ""
  )
}
```
